$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-02 paper/paper.tex
Bring in Erika's suggestions

% Functions
\newcommand{\subst}[3]{#1 \; [ \; #2 \; / \;#3 \;]}
\newcommand{\opsquared}[3]{#1 \, \coloneqq \, #2 \, \cdot^2 \, #3}
\newcommand{\opctx}[3]{#1 \, \coloneqq \, #2 \, \otimes \, #3}

@@

% Types
\newcommand{\Set}{\type{SET}}
\newcommand{\reduce}[1]{\; \type{\longrightarrow}_{#1} \;}
\newcommand{\types}[4]{#1 \; \type{\propto} \; #2 \; \type{\vdash} \; #3 \; \type{\boxtimes} \; #4}
\newcommand{\contains}[6]{#1 \; \type{\propto} \; #2 \; \type{\ni}_{#3} \; #4 \; \type{\propto} \; #5 \; \type{\boxtimes} \; #6}
\newcommand{\containsusage}[4]{#1 \; \type{\ni}_{#2} \; #3 \; \type{\boxtimes} \; #4}
\newcommand{\Var}{\type{VAR}}
\newcommand{\Process}{\type{PROCESS}}
\newcommand{\Unused}{\type{UNUSED}}
\newcommand{\PreCtx}{\type{PRECTX}}
\newcommand{\Ctx}{\type{CTX}}
\newcommand{\Type}{\type{TYPE}\;}
\newcommand{\Idx}{\type{IDX}\;}
\newcommand{\Idxs}{\type{IDXS}}
\newcommand{\Carrier}{\type{CARRIER}}
\newcommand{\N}{\type{\mathbb{N}}}
\newcommand{\Channel}{\type{CHANNEL}}
\newcommand{\Rec}{\type{REC}}
\newcommand{\Quantifier}{\type{QUANTIFIER}}
\newcommand{\eq}[1]{\; \type{=}_{#1} \;}
\newcommand{\eqeq}{\; \type{\equiv} \;}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{$\pi$ with leftovers: \\ a mechanisation in Agda}
\titlerunning{$\pi$ with leftovers: all mechanisation in Agda}
\author{Uma Zalakain}{University of Glasgow, Scotland}
       {u.zalakain.1@research.gla.ac.uk}{https://orcid.org/0000-0002-3268-9338}{}
\author{Ornela Dardha}{University of Glasgow, Scotland}
       {ornela.dardha@glasgow.ac.uk}{https://orcid.org/0000-0001-9927-7875}{}
\authorrunning{U. Zalakain and O. Dardha}
\Copyright{Uma Zalakain and Ornela Dardha}
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003753.10003761.10003764</concept_id>
<concept_desc>Theory of computation~Process calculi</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[300]{Theory of computation~Process calculi}
\keywords{pi-calculus, linear types, leftover typing, concurrency, mechanization, Agda}
\supplement{\url{https://github/umazalakain/typing-linear-pi}}
\acknowledgements{We want to thank Wen Kokke, James Wood, Guillaume Allais, Bob Atkey, and Conor McBride for their thoughts, patience, work, education and camaraderie.}
\funding{Supported by the UK EPSRC grant EP/K034413/1, ``From Data Types to Session Types: A Basis for Concurrency and Distribution'' (ABCD), and by the EU HORIZON 2020 MSCA RISE project 778233
  ``Behavioural Application Program Interfaces'' (BehAPI).}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\todo{rename Scoped to Process in code}
\todo{make sure all citations are there}
\todo{change colors, color functions}
\todo{do not distinguish fields from functions, it is not worth it}
\todo{go over guillaume's paper again}

@@

We keep track of the channel over which communication occurs so that we can state subject reduction.
We use a $dec$ function to decrement the index of this channel every time we come out of a binder.
This function saturates at $\constr{internal}$ --- meaning communication happens internally to the process.

@@

  \datatype
  {i : \Channel_n \\ P \; Q : \Process_n}
  {P \reduce{i} Q : \Set}
  \; \textsc{Reduces}

@@

  \inferrule
  {red : P \reduce{i} P'}
  {\constr{par} \; red : \comp{P}{Q} \reduce{i} \comp{P'}{Q}}

  \inferrule
  {red : P \reduce{i} Q}
  {\constr{res} \; red : \new P \reduce{dec\; i} \new Q}

  \inferrule
  {eq : P = P' \\ red : P' \reduce{i} Q}
  {\constr{struct} \; eq \; red : P \reduce{i} Q}
\end{mathpar}
\caption{Operational semantics indexed by the channel over which reduction occurs.}
\label{fig:reduction}
\end{figure}

@@

\begin{note}
  We will often work with pairs of input and output multiplicities.
  We use the notation $\Carrier^2$ to stand for a $\Carrier \times \Carrier$ tuple of input and output multiplicities, respectively.
  We use $\op{x}{y}{z}$ to stand for a monoidal operation on pairs of multiplicities -- where the algebraic laws are trivially lifted.
  Henceworth, we use $\lz$ to denote the pair $(0 \comma 0)$, $\li$ for the pair $(1 \comma 0)$, $\lo$ for $(0 \comma 1)$ and $\lio$ for $(1 \comma 1)$, borrowing from the linear \picalc{} notation \cite{KPT96,Sangio01}.
\end{note}

@@

\begin{description}
  \item [Linear]
    The carrier is a type $\type{Two}$ with the trivial elements $\constr{zero}$ and $\constr{one}$.
    The monoidal operation has the element $\constr{zero}$ as neutral on both sides, and the element $\constr{one}$ splitting into $\constr{one}$ and $\constr{zero}$ (or $\constr{zero}$ and $\constr{\one}$), and is uninhabited in every other case.
    All other rules follow trivially.

@@

\begin{nidefinition}
  \autoref{fig:prectx} defines types as either unit types, base types or channel types.
  The unit type $\unit$ serves as a proof of inhabitance for types.
  The base type $\base{n}$ uses natural numbers as placeholders for types --- the host language can interpret the former into the latter.
  This is for convenience, so that there is less universe polymorphism poluting specifications.
  The type $\channel{t}{x}$ of a channel determines what type of data and what usage annotations are sent over that channel.
  We keep track of types in a length-indexed \emph{pre-context} $\PreCtx_n$.

@@

    \inferrule
    {t : \Type \\ \stacked{idx : \Idx \\\\ x : \Carrier_{idx}}}
    {\channel{t}{x} : \Type}
  \end{mathpar}

@@

Framing captures the idea that the well-typedness of a process does not depend on its unused resources.
These unused resources are not introduced as new variables -- the typing context remains unaltered -- but rather as extra multiplicities that are carried from the input context to the output context.

>>>>

% Functions
\newcommand{\subst}[3]{#1 \; [ \; #3 \; \mapsto \;#2 \;]}
\newcommand{\opsquared}[3]{#1 \, \coloneqq \, #2 \, \cdot^2 \, #3}
\newcommand{\opctx}[3]{#1 \, \coloneqq \, #2 \, \otimes \, #3}

@@

% Types
\newcommand{\Set}{\type{SET}}
\newcommand{\reduce}[1]{\; \type{\longrightarrow}_{#1} \;}
\newcommand{\types}[4]{#1 \; \type{\propto} \; #2 \; \type{\vdash} \; #3 \; \type{\boxtimes} \; #4}
\newcommand{\contains}[6]{#1 \; \type{\propto} \; #2 \; \type{\ni}_{#3} \; #4 \; \type{\propto} \; #5 \; \type{\boxtimes} \; #6}
\newcommand{\containsusage}[4]{#1 \; \type{\ni}_{#2} \; #3 \; \type{\boxtimes} \; #4}
\newcommand{\Var}{\type{VAR}}
\newcommand{\Process}{\type{PROCESS}}
\newcommand{\Unused}{\type{UNUSED}}
\newcommand{\PreCtx}{\type{PRECTX}}
\newcommand{\Ctx}{\type{CTX}}
\newcommand{\Type}{\type{TYPE}\;}
\newcommand{\Idx}{\type{IDX}\;}
\newcommand{\Idxs}{\type{IDXS}}
\newcommand{\Carrier}{\type{CARRIER}}
\newcommand{\N}{\type{\mathbb{N}}}
\newcommand{\Channel}{\type{CHANNEL}}
\newcommand{\Rec}{\type{REC}}
\newcommand{\Quantifier}{\type{QUANTIFIER}}
\newcommand{\eq}[1]{\; \type{=}_{#1} \;}
\newcommand{\eqeq}{\; \type{\equiv} \;}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{$\pi$ with leftovers: \\ a mechanisation in Agda}
\titlerunning{$\pi$ with leftovers: all mechanisation in Agda}
\author{Uma Zalakain}{University of Glasgow, Scotland}
       {u.zalakain.1@research.gla.ac.uk}{https://orcid.org/0000-0002-3268-9338}{}
\author{Ornela Dardha}{University of Glasgow, Scotland}
       {ornela.dardha@glasgow.ac.uk}{https://orcid.org/0000-0001-9927-7875}{}
\authorrunning{U. Zalakain and O. Dardha}
\Copyright{Uma Zalakain and Ornela Dardha}
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003753.10003761.10003764</concept_id>
<concept_desc>Theory of computation~Process calculi</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[300]{Theory of computation~Process calculi}
\keywords{pi-calculus, linear types, leftover typing, concurrency, mechanization, Agda}
\supplement{\url{https://github/umazalakain/typing-linear-pi}}
\acknowledgements{We want to thank Erika Kreuter, Wen Kokke, James Wood, Guillaume Allais, Bob Atkey, and Conor McBride for their thoughts, patience, work, education and camaraderie.}
\funding{Supported by the UK EPSRC grant EP/K034413/1, ``From Data Types to Session Types: A Basis for Concurrency and Distribution'' (ABCD), and by the EU HORIZON 2020 MSCA RISE project 778233
  ``Behavioural Application Program Interfaces'' (BehAPI).}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\todo{make sure all citations are there}
\todo{change colors, color functions}
\todo{do not distinguish fields from functions, it is not worth it}
\todo{go over guillaume's paper again}

@@

We keep track of the channel over which communication occurs so that we can state subject reduction.
We use a $dec$ function to decrement the index of this channel every time we come out of a binder.
This function saturates at $\constr{internal}$: wrapping an internally reducing process with a binder will result in an internally reducing process.

@@

  \datatype
  {c : \Channel_n \\ P \; Q : \Process_n}
  {P \reduce{c} Q : \Set}
  \; \textsc{Reduces}

@@

  \inferrule
  {red : P \reduce{c} P'}
  {\constr{par} \; red : \comp{P}{Q} \reduce{c} \comp{P'}{Q}}

  \inferrule
  {red : P \reduce{c} Q}
  {\constr{res} \; red : \new P \reduce{dec\; c} \new Q}

  \inferrule
  {eq : P = P' \\ red : P' \reduce{c} Q}
  {\constr{struct} \; eq \; red : P \reduce{c} Q}
\end{mathpar}
\caption{Operational semantics indexed by the channel over which reduction occurs.}
\label{fig:reduction}
\end{figure}

@@

\begin{note}
  We will often work with pairs of input and output multiplicities.
  We use the notation $\Carrier^2$ to stand for a $\Carrier \times \Carrier$ tuple of input and output multiplicities, respectively.
  We use $\op{x}{y}{z}$ to stand for a monoidal operation on pairs of multiplicities -- where the algebraic laws are lifted elementwise.
  Henceworth, we use $\lz$ to denote the pair $(0 \comma 0)$, $\li$ for the pair $(1 \comma 0)$, $\lo$ for $(0 \comma 1)$ and $\lio$ for $(1 \comma 1)$, borrowing from the linear \picalc{} notation \cite{KPT96,Sangio01}.
\end{note}

@@

\begin{description}
  \item [Linear]
    The carrier is a type $\type{Two}$ with the trivial elements $\constr{zero}$ and $\constr{one}$.
    The monoidal operation has the element $\constr{zero}$ as neutral on both sides, and the element $\constr{one}$ splitting into $\constr{one}$ and $\constr{zero}$ (or $\constr{zero}$ and $\constr{one}$), and is uninhabited in every other case.
    All other properties are lifted elementwise.

@@

\begin{nidefinition}
  \autoref{fig:prectx} defines types as either unit types, base types or channel types.
  The unit type $\unit$ serves as a proof of inhabitance for types.
  The base type $\base{n}$ uses natural numbers as placeholders for types --- the host language can interpret the former into the latter.
  This is for convenience, so that there is less universe polymorphism polluting specifications.
  The type $\channel{t}{x}$ of a channel determines what type of data and what usage annotations are sent over that channel.
  We keep track of types in a length-indexed \emph{pre-context} $\PreCtx_n$.

@@

    \inferrule
    {t : \Type \\ \stacked{idx : \Idx \\\\ x : \Carrier_{idx}^2}}
    {\channel{t}{x} : \Type}
  \end{mathpar}

@@

Framing captures the idea that the well-typedness of a process only depends on the resources consumed by it.
These unused resources are not introduced as new variables -- the typing context remains unaltered -- but rather as extra multiplicities that are carried from the input context to the output context.
